---
title: "Heteroscedasticity Practical in R"
author: |
  | Bidyut Kumar Ghosh
  | Department of Commerce
  | Manipal Academy of Higher Education
  | email: bidyut.ghosh@manipal.edu
date: "`r format(Sys.time(), '%A, %B %d, %Y, %X')`"
output: slidy_presentation

editor_options: 
  chunk_output_type: inline
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment = NA, warning = FALSE,message = FALSE)
```

# Data Description

```{r, echo=FALSE}
htmltools::img(src = "logo1.png", 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;')
```

The file 'houseprice.xls' contains house price data from of a sample of 88 London houses together with some of their characteristics. 

The variables are:

1. Price = the price of the houses measured in pounds
2. Rooms = the number of bedrooms in each house
3. Sqfeet = the size of the house measured in square feet

We want to see whether the number of bedrooms and the size of the house play an important role in determining the price of each house.

$$ price=\beta_1+\beta_2 rooms+\beta_3 sqfeet+u $$

# Data Import

```{r, echo=FALSE}
htmltools::img(src = "logo1.png", 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;')
```



```{r}
library(readxl)
houseprice<-read_excel("houseprice.xls",range = cell_cols("A:J"))
head(houseprice)
```

# Data Visualisation

```{r, echo=FALSE}
htmltools::img(src = "logo1.png", 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;')
```



```{r}
library(ggplot2)
ggplot(houseprice,aes(sqrft,price))+geom_point(col="red")
```

```{r}
ggplot(houseprice,aes(bdrms,price))+geom_point(col="red")
```

# Model Estimation


```{r, echo=FALSE}
htmltools::img(src = "logo1.png", 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;')
```



```{r}
model1<-lm(price~sqrft+bdrms,data = houseprice)
summary(model1)
```

# Diagnostic Checking Using Plot function

```{r, echo=FALSE}
htmltools::img(src = "logo1.png", 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;')
```



```{r}
plot(model1)
```

# Formal Tests for Heteroscedasticity

- Breusch-Pagan test

```{r, echo=FALSE}
htmltools::img(src = "logo1.png", 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;')
```



```{r}
library(lmtest)
bptest(model1,  studentize = F)
```
- White's test


```{r}
bptest(model1,~bdrms*sqrft+I(bdrms^2)+I(sqrft^2), data = houseprice,studentize = F)
```

# Solving Heteroscedasticity (White's Method)

```{r, echo=FALSE}
htmltools::img(src = "logo1.png", 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;')
```



```{r}
library(sandwich)
coeftest(model1, vcov. = vcovHC(model1, type = "HC0"))
```

```{r}
coeftest(model1)
```


# Sovle for Heteroscedasticity (weighted least squares)

```{r, echo=FALSE}
htmltools::img(src = "logo1.png", 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px;')
```


Assuming that the variable causing the heteroskedasticity is the $sqfeet$ variable (or in mathematical notation assuming that):

$$ Var(u_i) = σ_i^2 = σ^ 2* sqfeet $$

Then we can use $\frac {1}{\sqrt {sqfeet}}$ as the weight.



```{r}
model2<-lm(price~sqrft+bdrms, data = houseprice, weights = 1/sqrt(sqrft))
summary(model2)
```

```{r}
plot(model2)
```


# Solve for Heteroscedasticity (by Log transformation)


$$  log(price)=\beta_1+\beta_2 rooms+\beta_3 log(sqfeet)+u  $$


```{r}
model3<-lm(log(price)~bdrms+log(sqrft),data = houseprice)
summary(model3)
```

```{r}
plot(model3)
```

```{r}
library(lmtest)
bptest(model3)
```

```{r}
bptest(model3,~bdrms*sqrft+I(bdrms^2)+I(sqrft^2), data = houseprice)
```

