---
author: "Moh"
date: "2023-08-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ANOVA FOR SPLIT-PLOT DESIGN
# Dataset 1: oats from MASS package in R
The yield of oats from a split-plot field trial using three varieties and four levels of manurial treatment. The experiment was laid out in 6 blocks of 3 main plots, each split into 4 sub-plots. The varieties were applied to the main plots and the manurial treatments to the sub-plots.

* A whole-plot is given by a plot of land in a block.
* The whole-plot factor is variety.
* A block design (RCB) was used at the whole-plot level.
* A split plot is given by a subplot of land.
* The split-plot factor is given by nitrogen treatment


```{r}
library(MASS)
oats<-data.frame(oats)
str(oats)
View(oats)
```

```{r}
summary(oats)
```

```{r}
sd(oats$Y)
```
```{r}
IQR(oats$Y)
```

```{r}
attach(oats)
boxplot(Y~B)


```

```{r}
boxplot(Y~V)
```


```{r}
boxplot(Y~B+V)
```

# In R we use the lmer function with an extra random effect (error) per 
# combination of block and variety.
```{r}
library(lme4) # for split-plot ANOVA
library(afex) # for p-values
fit.lme<-lmer(Y~B+V*N+(1|B:V), data=oats)
anova(fit.lme)

```

# R TUTORIAL: ANOVA FOR CROSS-OVER DESIGN

```{r}
library(nlme)
data1<-data.frame(read.csv("COD2.csv", header = TRUE))
attach(data1)
str(data1)

```


```{r}
summary(data1)
```


```{r}
boxplot(response~treatment)
```


```{r}
boxplot(response~period)
```


```{r}
boxplot(response~seq)
```

```{r}
L2 <- lme(response ~ period + seq + treatment,random=~1|sub_id,data=data1)
summary(L2)
```


```{r}
anova(L2)
```


```{r}

```

