---
title: "Factorial Design"
author: "Moh"
date: "2023-07-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#FACTORIAL DESIGNS IN R: TUTORIAL

# Note: Package: “car”, “emmeans” to be installed prior to running the 
# following code.

# Case 2: Tooth Growth in R
# call an in-built dataset in R

The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose 
levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery 
methods, orange juice or ascorbic acid (a form of vitamin C and coded 
as VC).

A 2×3 factorial design is a type of experimental design that allows 
researchers to understand the effects of two independent variables on a single dependent variable. In this type of design, one independent 
variable has two levels and the other independent variable has three 
levels.

```{r}
attach(ToothGrowth)
ToothGrowth

```
```{r}
help("ToothGrowth")
```

```{r}
str(ToothGrowth)
```

```{r}
require(graphics)
coplot(len ~ dose | supp, data = ToothGrowth, panel = panel.smooth,
       xlab = "ToothGrowth data: length vs dose, given type of supplement")
```

```{r}
summary(ToothGrowth)
```
```{r}
ToothGrowth$dose = factor(ToothGrowth$dose, levels=c(0.5,1.0,2.0), labels=c("low","med","high"))
str(ToothGrowth)

```


```{r}
table(ToothGrowth$supp, ToothGrowth$dose)
```

```{r}
boxplot(len ~ supp, data = ToothGrowth, ylab = 'Tooth Length', main = 'Box plot of Tooth Growth Data', col = 'steelblue')
```

```{r}
boxplot(len ~ dose, data=ToothGrowth, ylab="Tooth Length", main="Boxplots of Tooth Growth Data")
```

```{r}
boxplot(len ~ supp * dose, data = ToothGrowth, ylab = 'Tooth Length', main = 'Boxplots of Tooth Growth Data')
```

```{r}
with(ToothGrowth,interaction.plot(x.factor=dose, trace.factor=supp, response=len, fun=mean, type="b",legend=T, ylab="Tooth Length", main="Interaction Plot",pch=c(1,19)))
# An interaction plot that helps see change in dose accross two levels of supplement withb the 
# response variable beign length. The dot represents  mean values
```
There seems to be no difference between supp at high dose
There seems to be a main effect of dose – higher dose results in 
higher tooth length
There doesn’t seem to be much of a main effect of supp – there is 
less difference between the 2 groups overall.

```{r}
# Normality Assumption Check. Try some statistical test as assignment.
hist(ToothGrowth$len)
```


```{r}
densityplot(ToothGrowth$len)
```
```{r}
library(e1071)
library(ggplot2)
library(graphics)
library(lattice)
require(stats)
```
```{r}
with(ToothGrowth, tapply(len, list(supp,dose), mean))
```
```{r}
with(ToothGrowth, tapply(len, list(supp,dose), sd))
```
# Use Type I Sums of Squares are Sequential, so the order of variables in the models makes a difference.
# Use Type II when there is no interaction, and
# Use Type III when there is interaction

# Suggested Reading: https://towardsdatascience.com/anovas-three-types-of-estimating-sums-of-squares-don-t-make-the-wrong-choice-91107c77a27a

```{r}
# Homogeneity of Variance Assumption check

# options(show.signif.stars=F)
 bartlett.test(len ~ interaction(supp,dose), data=ToothGrowth)

```
```{r}

```

```{r}
bartlett.test(len~as.factor(supp), data=ToothGrowth)
```

```{r}
bartlett.test(len~as.factor(dose), data=ToothGrowth)
```

```{r}
library(car)
library(emmeans)
```
```{r}
fit = lm(len = supp + dose + supp : dose, data = ToothGrowth)
# fit = lm(len ~ supp+dose+supp*dose, data=ToothGrowth)
```
```{r}
aov.out2 = Anova(fit, data=ToothGrowth,type=3)
aov.out2
```
```{r}
# options(show.signif.stars= F) to off stars in results
# options(show.signif.stars= T) to off stars
aov.out2 = Anova(lm(len ~ supp+dose+supp*dose), data=ToothGrowth,type=3)
aov.out2
```
```{r}
marginal = emmeans(fit,~dose)
```

```{r}
pairs(marginal,adjust="tukey")
```
```{r}
marginal = emmeans(fit,~ supp)
pairs(marginal,adjust="tukey")

```
```{r}
marginal = emmeans(fit,~ supp:dose)
pairs(marginal,adjust="tukey")

```
```{r}
fit
```
```{r}
summary(fit)
```

```{r}
gog <-as.data.frame(read.csv(file.choose()))
```


```{r}
attach(gog)
```


```{r}
table(gender,alcohol)
```

```{r}
gog$gender = factor(gog$gender, levels=c("Female", "Male"))
gog$alcohol= factor(gog$alcohol, levels=c("No", "Yes"))
str(gog)

```

```{r}
summary(gog)
```
```{r}
boxplot(attractiveness ~ gender * alcohol, data=gog, ylab="Attractiveness", main="Boxplots of Goggles Data")
```
```{r}
with(gog,interaction.plot(x.factor=gender, trace.factor=alcohol, response=attractiveness, fun=mean, type="b",legend=T, ylab="Attractiveness", main="Interaction Plot",pch=c(1,19)))
```
```{r}
bartlett.test(attractiveness ~ interaction(gender, alcohol), data=gog)
```
```{r}
fit = lm((attractiveness ~ gender+alcohol+gender*alcohol), data = gog)
```

```{r}
#aov.out2 = Anova(lm(attractiveness ~ gender+alcohol+gender*alcohol), 
#data= gog,type=3)
#aov.out2
aov.out2 = Anova(fit, data=gog,type=3)
aov.out2

```
```{r}
fit2 = lm(attractiveness ~ gender+alcohol+gender*alcohol)
```

```{r}
marginal = emmeans(fit2,~ gender:alcohol)
pairs(marginal,adjust="tukey")
```

